---
title: "Week_06_Genetic_Algorithm_Assignment"
author: "saikiran"
date: "2025-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
# Load required libraries
library(data.table)
library(geosphere)
library(GA)
library(purrr)
library(sf)
```

```{r}
# Load population data
population_data <- fread("Mo_pop_Sim.csv")
str(population_data)

# Sample 0.01% of the population data
sampling_fraction <- 0.0001
sample_population <- population_data[sample(1:nrow(population_data), size = round(nrow(population_data) * sampling_fraction), replace = FALSE), ]

```
```{r}
# Load Federally Qualified Health Center (FQHC) data
fqhc_data <- st_read("~/Desktop/week6Assigment/MO_2018_Federally_Qualified_Health_Center_Locations.shp")
fqhc_locations <- data.table(
  ID = 1:nrow(fqhc_data),
  Longitude = fqhc_data$Longitude,
  Latitude = fqhc_data$Latitude,
  Facility_Name = fqhc_data$Facility,
  Facility_Group = fqhc_data$Group_Name
)

```
```{r}
# Function to calculate average distance (accessibility) for each FQHC
calculate_average_distance <- function() {
  avg_distances <- numeric(nrow(fqhc_locations))

  for (i in 1:nrow(fqhc_locations)) {
    fqhc_coordinates <- c(fqhc_locations$Longitude[i], fqhc_locations$Latitude[i])
    distances <- distHaversine(
      matrix(c(sample_population$long, sample_population$lat), ncol = 2),
      matrix(rep(fqhc_coordinates, each = nrow(sample_population)), ncol = 2)
    )
    avg_distances[i] <- mean(distances)
  }
  
  fqhc_locations[, avg_distance := avg_distances]
  return(fqhc_locations)
}

```

```{r}
# Function to calculate population density within a 30-mile radius (48280 meters)
calculate_population_density <- function() {
  density_counts <- numeric(nrow(fqhc_locations))

  for (i in 1:nrow(fqhc_locations)) {
    fqhc_coordinates <- c(fqhc_locations$Longitude[i], fqhc_locations$Latitude[i])
    distances <- distHaversine(
      matrix(c(sample_population$long, sample_population$lat), ncol = 2),
      matrix(rep(fqhc_coordinates, each = nrow(sample_population)), ncol = 2)
    )
    density_counts[i] <- sum(distances <= 48280)  # Count points within 30 miles
  }

  fqhc_locations[, population_density := density_counts * (1 / sampling_fraction)]
  return(fqhc_locations)
}

```

```{r}
# Calculate the metrics for all FQHC locations
fqhc_locations <- calculate_average_distance()
fqhc_locations <- calculate_population_density()

```

```{r}
# Rank FQHCs by accessibility (average distance)
top_accessible <- fqhc_locations[order(avg_distance)][1:8]
print(top_accessible[, .(ID, Facility_Name, avg_distance)])

```

```{r}
# Rank FQHCs by population density
top_populated <- fqhc_locations[order(-population_density)][1:8]
print(top_populated[, .(ID, Facility_Name, Facility_Group, population_density)])

```

```{r}
# Combined ranking by both metrics: Access and Population Density
fqhc_locations[, access_rank := rank(avg_distance)]
fqhc_locations[, density_rank := rank(-population_density)]
fqhc_locations[, combined_rank := access_rank + density_rank]

top_combined_fqhc <- fqhc_locations[order(combined_rank)][1:8]
print(top_combined_fqhc[, .(ID, Facility_Name, Facility_Group, combined_rank, avg_distance, population_density)])

```

```{r}
# Genetic Algorithm: Fitness function
fitness_function <- function(selection) {
  selected_fqhc <- fqhc_locations[selection == 1, ]

  # Compute the total combined score for selected FQHCs
  total_combined_score <- sum(selected_fqhc$combined_rank)

  # We want to minimize the combined score
  return(-total_combined_score)
}

```

```{r}
# Run the Genetic Algorithm to optimize FQHC location selection
set.seed(123)
genetic_algorithm <- ga(
  type = "binary",
  fitness = fitness_function,
  nBits = nrow(fqhc_locations),
  popSize = 50,
  maxiter = 100,
  pmutation = 0.1,
  elitism = 2
)

```

```{r}
# Extract the best solution
best_solution <- as.logical(genetic_algorithm@solution[1, ])
optimal_fqhc <- fqhc_locations[best_solution, ]

# Print the optimal FQHC locations based on Genetic Algorithm optimization
print(optimal_fqhc[, .(ID, Facility_Name, Facility_Group, combined_rank, avg_distance, population_density)])

```

Conclusion

The genetic algorithm optimization through its evaluation method discovered eight Federally Qualified Health Centers (FQHCs) which proved suitable for specialized mental health service implementation. Accessibility and population density were the main factors in the selection process through which a fitness function determined facility placement. The function measured both aspects by examining residence distance and population distribution within 30 miles.

The selected FQHCs include COMTREA Byrnes Mill Health Center and Access Family Care - Anderson (Dental) which provide optimal service areas near residents and population concentration zones. The optimization strategy determines these specific locations because they maximize mental health service accessibility and population coverage throughout Missouri.
